Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-01-27 paper2.qmd
section:appendix; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the growing number of CJ studies, unsystematic and fragmented research approaches have left several critical issues unresolved. This study addresses two such issues that can compromise the reliability and validity of CJ's trait estimates. The first concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

<!-- 4. what is the organization of the study? -->
The study is organized into six main sections. It begins with an overview of Thurstone's theory in @sec-thurstone_theory, followed by a detailed examination of these issues in @sec-theory-issues. In @sec-theoretical, the study develops a theoretical model for CJ that simultaneously addresses these issues. The model integrates Thurstone's core principles with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a statistical model that supports the analysis of pairwise comparison data. Finally, @sec-discussion discusses the concepts developed in the study, their implications, and avenues for future exploration, while @sec-conclusion summarizes the findings.

@@

# The two critical issues in CJ literature {#sec-theory-issues}

This section explores the two critical issues in CJ's theory that have the potential to compromise the reliability and validity of CJ's trait estimates. @sec-theory-issue1 inspects the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a problem of trait scaling. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each case progressively incorporated additional simplifying assumptions into the model.

@@

In aggregate, researchers' inadequate handling of violations to the assumptions of equal dispersion and zero correlation between stimuli, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, can potentially compromise the reliability of the trait estimates and, by extension, their validity [@Perron_et_al_2015, pp. 2]. Consequently, adopting a more systematic and integrated approach to handling these assumptions and examining the factors influencing CJ experiments could offer several statistical and measurement benefits, including the ability to address the aforementioned issues.

@@

## The population model {#sec-theory-theoretical_P}

@@

<!-- 2. What do they do next? -->
Ideally, researchers then proceed to identify the estimands. *Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Identification is crucial because if an estimator cannot reliably compute an estimand using "infinite" and error-free data, it is impossible to gain meaningful insights about that estimand using finite data [@Schuessler_et_al_2023, pp. 5]. An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand derived through the process of *estimation*, which integrates data with an estimator [@Everitt_et_al_2010]. The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], shown in @fig-IEflow, visually represents the transition from estimands to estimates.

@@

<!-- 3. But, what do we need to produce a good estimate? -->
Consequently, although many estimators can approximate an estimand, researchers prioritize those with desirable identification properties to ensure accurate and reliable estimates. For instance, the Z-test is a widely used estimator for comparing group proportions, providing accurate estimates when its underlying assumptions are satisfied [@Kanji_2006]. Moreover, researchers can interpret the Z-test estimates as causal, provided they collect the data through a randomized experiment.

@@

<!-- 3. But how does it work? -->
Even though counterfactuals are unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach. The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones. When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37]. Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp. 32] (see @eq-group_diff).

@@

For instance, If researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

SCMs and DAGs support causal inference through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35].

@sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.

@@

<!-- 1. Transforming probabilistic association into causation?  -->
The structural approach to causal inference translate the probabilistic insights from @sec-appendix-B23 into actionable strategies seeking to identify the ACE from associational quantities. It achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp. 34].

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$.

Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE. This process is expressed in @eq-mACE2 (see @sec-appendix-B). Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.

@@

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming that variables exhibit specific local (in)dependencies. These assumptions enhance model tractability and facilitate the estimation process. As described in @sec-appendix-B22, researchers can use SCMs and DAGs to formalize these local (in)dependencies, which leads to the BNF of the likelihood (@eq-like_BNF). From this point, any probabilistic programming language can model this simpler structure and generate the parameter's posterior distribution using @eq-bayes_rule.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of studies on CJ, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study addresses two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates. The first problem concerns the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. The second involves the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

<!-- 4. what is the organization of the study? -->
The study is organized into six main sections. An overview of Thurstone's theory is presented in @sec-thurstone_theory, followed by a detailed examination of the identified issues in @sec-theory-issues. In @sec-theoretical, the study proposes a theoretical model for CJ that addresses these challenges. The model incorporates Thurstone's core principles alongside key assessment design features relevant to CJ experiments, such as judges, stimuli, and comparison selections. In @sec-statistical, these practical and theoretical considerations are translated into a statistical model for the analysis of pairwise comparison data. Finally, @sec-discussion explores the implications of the study's findings and the avenues for future exploration, and @sec-conclusion provides a summary of the key insights.

@@

# Two prominent issues in CJ literature {#sec-theory-issues}

This section discusses the two prominent issues in CJ literature that can undermine the reliability and validity of CJ's trait estimates. @sec-theory-issue1 examines the overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data. @sec-theory-issue2 focuses on the apparent disconnect between CJ's approach to trait measurement and hypothesis testing.

@@

<!-- 1. What is the problem with the general form -->
Thurstone noted from the outset that the general form of the theory, as outlined in @sec-thurstone_theory, gave rise to a trait scaling problem. Specifically, the model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical implementation of the theory, he developed five cases derived from this general form, each progressively incorporating additional simplifying assumptions into the model.

@@

<!-- 4. segway for next section -->
In aggregate, researchers' overreliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates. This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp. 2]. Consequently, addressing these issues and examining the factors influencing CJ experiments by adopting a more systematic and integrated approach could offer several statistical and measurement benefits.

@@

## The conceptual population model {#sec-theory-theoretical_P}

@@

<!-- 2. What do they do next? -->
Researchers then proceed to identify the estimands. *Identification* determines whether an estimator can accurately compute the estimand based solely on its assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. An *estimator* refers to a method or function that transforms data into an estimate [@Neal_2020]. *Estimates* are numerical values that approximate the estimand derived through the process of *estimation*, which integrates data with an estimator [@Everitt_et_al_2010]. The Identification-Estimation flowchart [@McElreath_2020; @Neal_2020], shown in @fig-IEflow, visually represents the transition from estimands to estimates.

@@

<!-- 3. the importance of identification -->
Identification is essential because it is a necessary condition for *consistency*. An estimator is *consistent* if it converges to the "true" value of an estimand as the sample size approaches infinity [@Everitt_et_al_2010]. Without identification, an estimator cannot achieve consistency, rendering it incapable of reliably computing the estimand, even with "infinite" and error-free data. Consequently, deriving meaningful insights about the estimand from finite data becomes impossible [@Schuessler_et_al_2023, pp. 5]. Therefore, to ensure accurate and reliable estimates, researchers prioritize estimators with desirable identification properties. For instance, the Z-test is a widely used estimator for comparing group proportions, yielding accurate estimates when its underlying assumptions are satisfied [@Kanji_2006]. Furthermore, researchers can interpret estimates from the Z-test as causal, provided the data is collected through a randomized experiment.

@@

<!-- 3. But how does it work? -->
Even though counterfactuals are unobservable, researchers can still identify the ACE from associational estimates by leveraging the structural approach. The approach identifies the ACE by statistically conditioning data on a *sufficient adjustment set* of variables $X$ [@Pearl_2009; @Pearl_et_al_2016; @Morgan_et_al_2014]. This *sufficient* set (potentially empty) must block all non-causal paths between $T$ to $Y$ without opening new ones. When such a set exists, then $T$ and $Y$ are *d-separated* by $X$ ($T \dsep Y \cond X$) [@Pearl_2009], and $X$ satisfies the *backdoor criterion* [@Neal_2020, pp 37]. Here, *conditioning* describes the process of restricting the focus to the subset of the population defined by the conditioning variable [@Neal_2020, pp. 32] (see @eq-CACE).

@@

For instance, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

SCMs and DAGs support causal inference through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35]. @sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.

@@

<!-- 1. Transforming probabilistic association into causation?  -->
The structural approach to causal inference translates probabilistic insights into actionable strategies seeking to identify the ACE from associational quantities. The approach achieves this by relying on the *modularity assumption*, which posits that intervening on a node alters only the causal mechanism of that node, leaving others unchanged [@Neal_2020, pp. 34].

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$. Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.

@@

Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE. This process is expressed in @eq-mACE2 (see @sec-appendix-B).

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables. SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendix-B22. These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF). With this simplified structure, any probabilistic programming language can model the system and compute the parameter's posterior distribution using @eq-bayes_rule.
